|

### 2019-09-02 paper.Rmd
Thickness for spheres

---
title: |
  A slice tour for finding hollowness in high-dimensional data
type: Short Technical Note
author:
  - name: Ursula Laa
    affil: a, b
    email: ursula.laa@monash.edu
  - name: Dianne Cook
    affil: b
    email: dicook@monash.edu
affiliation:
  - num: a
    address: |
      School of Physics and Astronomy, Monash University
  - num: b
    address: |
      Department of Econometrics and Business Statistics, Monash University
bibliography: biblio.bib
abstract: |
  Projections and sectioning are complementary methods for the visualisation of
  high-dimensional data. In this note we describe slicing in the orthogonal
  space of projections obtained when running a tour, thus presenting the viewer
  with an interpolated sequence of projected sections. The method has been
  implemented in R as an extension to the tourr package, and can be used to
  explore concave structures in multivariate distributions.
keywords: |
  data visualisation; grand tour; sectioning
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
output: rticles::tf_article
---

@@

The distance vector can then be used to display a slice tour by highlighting points for which the orthogonal distance is smaller than a selected cutoff value $h$. In general a larger cutoff value and/or much larger sample sizes are required as the original dimensionality ($p$) increases. For practical purposes we therefore specify a volume $\epsilon$ and then use as a cutoff $h = \epsilon^{1/(p-2)}$. XXX not yet implemented that way, based on thickness estimates

@@

Choosing the slice thickness is a compromise between what feature size can be revealed and how many points fall inside a thin slice. As dimensionality increases, the relative number of points that are inside a slice of fixed thickness will decrease. The exact relation depends heavily on the distribution of the data points. Here we present expressions for spectial cases to get an idea of this scaling.

We first consider as a simple example a hypercube that is sliced along two of the paremeters (note that a cube is not rotation invariant). To estimate the relative number of points in a slice of thickness $2h$ we calculate the relative volume of the slice compared to the full hypercube:
\begin{equation}
V_{rel} = \frac{V_{slice}}{V_{cube}} = \frac{(2r)^2 * (2h)^{p-2}}{(2r)^p} = (\frac{h}{r})^{p-2},
\end{equation}
where the cube side lenght is $2r$. Thus if we want to keep the relative number of points (i.e. $V_{rel}$) constant for slices in different $p$ we calculate $h$ from a fixed volume parameter as $h = \epsilon^{1/(p-2)}$.

Considering now slicing through a multivariate normal, we can estimate the number of points in a slice by integrating the probability distribution with boundaries $\pm h$ in $p-2$. In this case the relative density scales with the error function as
\begin{equation}
\rho_{rel} = \pi^{(2-p)/2} \mathrm{erf}((h/\sqrt 2 \sigma)) ^{p-2}.
\end{equation}

>>>>

---
title: |
  A slice tour for finding hollowness in high-dimensional data
type: Short Technical Note
author:
  - name: Ursula Laa
    affil: a, b
    email: ursula.laa@monash.edu
  - name: Dianne Cook
    affil: b
    email: dicook@monash.edu
affiliation:
  - num: a
    address: |
      School of Physics and Astronomy, Monash University
  - num: b
    address: |
      Department of Econometrics and Business Statistics, Monash University
bibliography: biblio.bib
abstract: |
  Projections and sectioning are complementary methods for the visualisation of
  high-dimensional data. In this note we describe slicing in the orthogonal
  space of projections obtained when running a tour, thus presenting the viewer
  with an interpolated sequence of projected sections. The method has been
  implemented in R as an extension to the tourr package, and can be used to
  explore concave structures in multivariate distributions.
keywords: |
  data visualisation; grand tour; sectioning
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \usepackage{setspace}\doublespacing
output: rticles::tf_article
---

@@

The distance vector can then be used to display a slice tour by highlighting points for which the orthogonal distance is smaller than a selected cutoff value $h$. Notice that in 3d, where for each plane we have one well-defined orthogonal direction, this defines a flat slice of hight $2*h$, matching the common conception of a slice. When going beyond 3d, the codimension of the plane is $>1$ and using the euclidean distance results in rotation invariant slicing in the orthogonal space. As a consequence the "slices" are now spherical in the orthogonal subspace, with radius $h$.

@@

Choosing the slice thickness is a compromise between what feature size can be revealed and how many points fall inside a thin slice. As dimensionality $p$ increases, the relative number of points that are inside a slice of fixed thickness $h$ will decrease. The exact relation depends heavily on the distribution of the data points. Here we present expressions for special cases to get an idea of this scaling. For practical purposes we specify a volume $\epsilon$ and then use as a cutoff $h = \epsilon^{1/(p-2)}$. XXX not yet implemented that way, based on thickness estimates.

We consider two types of rotation invariant distributions for our estimates: a hypersphere in $p$ dimensions and a $p$ dimensional normal distribution without correlation.

Let us consider points uniformly distributed in a $p$-dimensional hypersphere. The fraction of points inside a slice of thickness $2*h$ can therefore be estimated as the relative volume of the slice compared to the volume of the full hypersphere,
\begin{equation}
V_{rel} = \frac{1}{2} \frac{h^{p-2}}{r^p} (p r^2 - (p-2) h^2) \approx \frac{1}{2} (\frac{h}{R})^{p-2},
\end{equation}
where the approximation is valid when $h \ll r$, as is typically expected to be the case. Thus if we want to keep the relative number of points (i.e. $V_{rel}$) approximately constant for slices in different dimensions $p$, we calculate $h$ from a volume parameter as $h = \epsilon^{1/(p-2)}$. XXXX change code to this? currently $\epsilon^{1/p}$ is implemented...

XXXX need to redo integration of multivariate normal...
Considering now slicing through a multivariate normal, we can estimate the number of points in a slice by integrating the probability distribution with boundaries $\pm h$ in $p-2$. In this case the relative density scales with the error function as
\begin{equation}
\rho_{rel} = \pi^{(2-p)/2} \mathrm{erf}((h/\sqrt 2 \sigma)) ^{p-2}.
\end{equation}
